---
title: "Cold-Related Deaths"
subtitle: "This indicator presents data on deaths classified as “cold-related” in the United States."
---

```{r}
#| label: setup
#| include: false
source(here::here("_common.R"))
source(here::here("R/figures.R"))
meta <- read_meta()
```

```{r}
#| label: caption-block
#| include: false
# Renders a figure's caption strip from data/meta.yml: title, data source, and
# web-update date, all copied verbatim from what R/build_data.R recorded.
# Nothing here is hand-typed, so it cannot drift from the source.
caption_block <- function(file) {
  m <- meta_for(file, meta)
  knitr::asis_output(sprintf(
    "**%s**\n\nData source: %s <br> Web update: %s\n",
    m$figure_title, m$data_source, m$web_update
  ))
}
```

## Figure 1

```{r}
#| label: fig-1-caption
#| echo: false
#| output: asis
caption_block("cold_deaths_annual.csv")
```

```{r}
#| label: fig-1
#| echo: false
fig_1()
```

This figure shows the annual rates for deaths classified as "cold-related" by
medical professionals in the 50 states and the District of Columbia. The blue
line shows deaths for which cold was listed as the main (underlying) cause.
The orange line shows deaths for which cold was listed as either the
underlying or contributing cause of death, based on a broader set of data
that became available in 1999.

Between 1998 and 1999, the World Health Organization revised the
international codes used to classify causes of death. As a result, data from
earlier than 1999 cannot easily be compared with data from 1999 and later.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-1-table
#| echo: false
knitr::kable(fig_1_table(), digits = 3)
```

Download: [tidy CSV](data/cold_deaths_annual.csv) ·
[original workbook](data-raw/cold-deaths_figure-1-and-TD1_04-08-19.xlsx)

</details>

## Figure TD-1

::: {.callout-note appearance="simple" icon=false}
This figure is not part of EPA's published indicator page. It is EPA's own
supplementary Figure TD-1 from the [technical
documentation](`r meta$indicator$technical_documentation`), built here
because the data exists and shows the seasonal pattern behind Figure 1's
underlying-or-contributing series (its blue line's 1999-2015 deaths,
aggregated by calendar month instead of by year). See
[Data & Downloads](data.qmd) and `data-raw/PROVENANCE.md` for details.
:::

```{r}
#| label: fig-td1-caption
#| echo: false
#| output: asis
caption_block("cold_deaths_monthly.csv")
```

```{r}
#| label: fig-td1
#| echo: false
fig_td1()
```

This graph shows all deaths for which cold was recorded as either the
underlying cause or a contributing cause of death, which corresponds with the
orange line in Figure 1 above. Deaths have been aggregated by month over the
entire period of record.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-td1-table
#| echo: false
knitr::kable(fig_td1_table())
```

Download: [tidy CSV](data/cold_deaths_monthly.csv) ·
[original workbook](data-raw/cold-deaths_figure-1-and-TD1_04-08-19.xlsx)

</details>

## Key Points

- Between 1979 and 2016, the death rate as a direct result of exposure to
  cold (underlying cause of death) generally ranged from 1 to 2.5 deaths per
  million people, with year-to-year fluctuations (see Figure 1). Overall, a
  total of more than 19,000 Americans have died from cold-related causes
  since 1979, according to death certificates.

- For years in which the two records overlap (1999–2015), accounting for
  those additional deaths in which cold was listed as a _contributing_
  factor results in a higher death rate—more than double for most years—compared
  with the estimate that only includes deaths where cold was listed
  as the _underlying_ cause (see Figure 1).

- While increases in deaths are generally associated with colder
  temperatures, some winter deaths are due to factors other than exposure to
  cold conditions. For example, winter is typically flu season. In other
  cases, even if cold exposure contributes to a death, it may not be
  reported as "cold-related" on a death certificate. These limitations, as
  well as year-to-year variability in the data and a change in
  classification codes in the late 1990s, make it difficult to determine
  whether the United States has experienced a meaningful increase or
  decrease in deaths classified as "cold-related" over time.

## Background

In recent years, U.S. death rates in winter months have been 8 to 12 percent
higher than in non-winter months.^[1](#ref-1)^ Much of this increase relates
to seasonal changes in behavior and the human body, as well as increased
exposure to respiratory diseases. Cold temperatures can also worsen
pre-existing medical conditions such as cardiovascular and respiratory
diseases. For example, death rates from heart attacks increase as
temperatures drop, likely due to the way cold affects blood circulation,
blood vessels, and other factors.^[2](#ref-2),[3](#ref-3)^ Even moderately
cold days can increase the risk of death for many people. People exposed to
extremely cold conditions can also suffer from direct effects such as
frostbite and potentially deadly hypothermia, especially in places where
people are not accustomed to cold temperatures.

Certain population groups face higher risks of cold-related illness or
death. For example, occupational groups that work outdoors during winter
months, such as agricultural workers, construction workers, and electricity
and pipeline utility workers, face higher risks of exposure to
cold.^[4](#ref-4)^ Others at risk include older adults, infants, people with
pre-existing medical conditions, people taking medications or using drugs
(especially alcohol) that make them more susceptible to cold effects,
homeless people, and those with inadequate winter clothing or home heating.^[5](#ref-5)^

Unusually cold winter temperatures have become less common across the
contiguous 48 states in recent decades,^[6](#ref-6)^ particularly very cold
nights (see the High and Low Temperatures indicator). Extreme cold waves are
likely to continue to decrease as winter temperatures increase in the
future.^[7](#ref-7)^ This winter warming is expected to reduce the number of
direct cold-related deaths, but the decrease is projected to be smaller than
increases in heat-related deaths (see the Heat-Related Deaths indicator) in
most regions.^[8](#ref-8)^ This is because some of the factors that lead to
higher death rates in the winter are not particularly sensitive to climate
change,^[9](#ref-9)^ because extreme heat has a more immediate and direct
effect on death rates than extreme cold,^[10](#ref-10)^ and because the
solutions to protect against cold exposure (such as staying indoors, wearing
more clothing, turning on the heat) are more widely accessible than
protection against extreme heat.^[11](#ref-11)^ Cold-related death rates can
change as communities strengthen their cold weather plans and take other
steps to protect vulnerable people during cold winter months.

## About the Indicator

Like to the Heat-Related Deaths indicator, this indicator shows the annual rate
for deaths classified by medical professionals as "cold-related" in the
United States based on death certificate records. Every death is recorded on
a death certificate, where a medical professional identifies the main cause
of death (also known as the underlying cause), along with other conditions
that contributed to the death. These causes are classified using a set of
standard codes. Dividing the annual number of deaths by the U.S. population
in that year, then multiplying by one million, will result in the death
rates (per million people) that this indicator shows.

Figure 1 shows cold-related death rates using two methods. One method shows
deaths for which exposure to excessive natural cold was stated as the
underlying cause of death from 1979 to 2016. The other data series shows
deaths for which cold was listed as either the underlying cause or a
contributing cause, based on a broader set of data that, at present, can
only be evaluated for the years 1999 to 2015. For example, in a case where
cardiovascular disease was determined to be the underlying cause of death,
cold could be listed as a contributing factor because it can make the
individual more susceptible to the effects of this disease.

## About the Data

### Indicator Notes

Several factors influence the ability of this indicator to estimate the true
number of deaths associated with exposure to cold temperatures. Some
cold-related deaths are not identified as such by the medical examiner and
might not be properly coded on the death certificate. In many cases, the
medical examiner might classify the cause of death as a cardiovascular or
respiratory disease, not knowing for certain whether cold was a contributing
factor. Furthermore, deaths can occur from exposure to cold (either as an
underlying cause or as a contributing factor) that is not classified as
extreme and therefore is often not recorded as such. These factors are
similar to the factors that likely lead to undercounting of heat-related
deaths (see the Heat-Related Deaths indicator), except that the effects of
cold on the body tend to persist for longer than the effects of
heat,^[14](#ref-14)^ which could make it even more difficult to connect
deaths to exposures.

Classifying a death as "cold-related" does not mean that cold temperatures
were the only factor that caused or contributed to the death, as
pre-existing medical conditions can significantly increase an individual's
susceptibility to cold temperatures. Other important factors, such as the
overall vulnerability of the population, and the extent to which people have
adapted and acclimated to cold temperatures, can also affect trends in
cold-related deaths. Some deaths classified as "cold-related" reflect human
behavior more than weather conditions—for instance, deaths in which alcohol
abuse increased a person's susceptibility to cold weather, falling into cold
water (which can cause hypothermia, even in non-winter months), or getting
lost on a wilderness excursion. This indicator includes these types of
deaths.

### Data Sources

Data for this indicator were provided by the U.S. Centers for Disease
Control and Prevention (CDC). The 1979–2016 underlying cause data in Figure
1 are publicly available through the CDC WONDER database at:
<https://wonder.cdc.gov/mortSQL.html>. The 1999–2015 analysis in
Figure 1 was developed by using data provided by CDC's Environmental Public
Health Tracking Program.

## Technical Documentation

```{r}
#| label: td-link
#| echo: false
#| output: asis
cat(sprintf("- [Download related technical information (PDF)](%s)\n", meta$indicator$technical_documentation))
```

## References

<ol class="references">
  <li id="ref-1">CDC (U.S. Centers for Disease Control and Prevention). 2017. CDC WONDER database: Compressed mortality file, underlying cause of death. Monthly all-cause mortality for 2011–2016. Accessed March 2018. <https://wonder.cdc.gov/mortSQL.html>.</li>
  <li id="ref-2">Gasparrini, A., et al. 2015. Mortality risk attributable to high and low ambient temperature: A multicountry observational study. Lancet 386(9991):369–375.</li>
  <li id="ref-3">Medina-Ramón, M., and J. Schwartz. 2007. Temperature, temperature extremes, and mortality: A study of acclimatization and effect modification in 50 U.S. cities. Occup. Environ. Med. 64(12):827–833.</li>
  <li id="ref-4">Sarofim, M.C., S. Saha, M.D. Hawkins, D.M. Mills, J. Hess, R. Horton, P. Kinney, J. Schwartz, and A. St. Juliana. 2016. Chapter 2: Temperature-related death and illness. In: The impacts of climate change on human health in the United States: A scientific assessment. U.S. Global Change Research Program. <https://health2016.globalchange.gov>.</li>
  <li id="ref-5">Berko, J., D.D. Ingram, S. Saha, and J.D. Parker. 2014. Deaths attributed to heat, cold, and other weather events in the United States, 2006–2010. National Health Statistics Reports No. 76. Hyattsville, Maryland: National Center for Health Statistics. <https://www.cdc.gov/nchs/data/nhsr/nhsr076.pdf>.</li>
  <li id="ref-6">USGCRP (U.S. Global Change Research Program). 2017. Climate science special report: Fourth National Climate Assessment, volume I. Wuebbles, D.J., D.W. Fahey, K.A. Hibbard, D.J. Dokken, B.C. Stewart, and T.K. Maycock (eds.). <https://science2017.globalchange.gov>. doi:10.7930/J0J964J6</li>
  <li id="ref-7">USGCRP (U.S. Global Change Research Program). 2017. Climate science special report: Fourth National Climate Assessment, volume I. Wuebbles, D.J., D.W. Fahey, K.A. Hibbard, D.J. Dokken, B.C. Stewart, and T.K. Maycock (eds.). <https://science2017.globalchange.gov>. doi:10.7930/J0J964J6</li>
  <li id="ref-8">Sarofim, M.C., S. Saha, M.D. Hawkins, D.M. Mills, J. Hess, R. Horton, P. Kinney, J. Schwartz, and A. St. Juliana. 2016. Chapter 2: Temperature-related death and illness. In: The impacts of climate change on human health in the United States: A scientific assessment. U.S. Global Change Research Program. <https://health2016.globalchange.gov>.</li>
  <li id="ref-9">Sarofim, M.C., S. Saha, M.D. Hawkins, D.M. Mills, J. Hess, R. Horton, P. Kinney, J. Schwartz, and A. St. Juliana. 2016. Chapter 2: Temperature-related death and illness. In: The impacts of climate change on human health in the United States: A scientific assessment. U.S. Global Change Research Program. <https://health2016.globalchange.gov>.</li>
  <li id="ref-10">Sarofim, M.C., S. Saha, M.D. Hawkins, D.M. Mills, J. Hess, R. Horton, P. Kinney, J. Schwartz, and A. St. Juliana. 2016. Chapter 2: Temperature-related death and illness. In: The impacts of climate change on human health in the United States: A scientific assessment. U.S. Global Change Research Program. <https://health2016.globalchange.gov>.</li>
  <li id="ref-11">Barnett, A.G., S. Hajat, A. Gasparrini, and J. Rocklöv. 2012. Cold and heat waves in the United States. Environ. Res. 112:218–224.</li>
  <li id="ref-12">CDC (U.S. Centers for Disease Control and Prevention). 2018. Mortality Data on CDC WONDER: Underlying cause-of-death: All ages deaths from compressed mortality. Accessed March 2018. <https://wonder.cdc.gov/mortSQL.html>.</li>
  <li id="ref-13">CDC (U.S. Centers for Disease Control and Prevention). 2018. Monthly national totals provided by National Center for Environmental Health staff in March 2018, based on contributing causes of death with exclusion criteria applied.</li>
  <li id="ref-14">Gasparrini, A., et al. 2015. Mortality risk attributable to high and low ambient temperature: A multicountry observational study. Lancet 386(9991):369–375.</li>
</ol>

::: {.callout-note appearance="simple" icon=false}
Text on this page is EPA's own published wording (a U.S. Government work, not
subject to domestic copyright), extracted from EPA's source document and
verified paragraph by paragraph against the published page. See
`R/gen_narrative.R` and `NOTICE.md`.
:::
