---
title: "TODO: Indicator Name"
subtitle: "TODO: EPA's one-line subtitle for this indicator."
---

```{r}
#| label: setup
#| include: false
source(here::here("_common.R"))
source(here::here("R/figures.R"))
meta <- read_meta()
```

```{r}
#| label: caption-block
#| include: false
# Renders a figure's caption strip from data/meta.yml: title, data source, and
# web-update date, all copied verbatim from the CSV EPA published. Nothing here
# is hand-typed, so it cannot drift from the source file.
caption_block <- function(file) {
  m <- meta_for(file, meta)
  knitr::asis_output(sprintf(
    "**%s**\n\nData source: %s <br> Web update: %s\n",
    m$figure_title, m$data_source, m$web_update
  ))
}
```

<!--
TODO: this indicator may have a different number of figures than
heat-related-deaths' three. Duplicate or delete the figure blocks below to
match. Each block's file name (e.g. "figure_1.csv") must match what
R/build_data.R writes to data/ and what data/meta.yml documents.
-->

## Figure 1

```{r}
#| label: fig-1-caption
#| echo: false
#| output: asis
caption_block("TODO_figure_1.csv")
```

```{r}
#| label: fig-1
#| echo: false
fig_1()
```

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-1-table
#| echo: false
knitr::kable(fig_1_table(), digits = 3)
```

Download: [tidy CSV](data/TODO_figure_1.csv) · [original EPA file](data-raw/TODO_source_1.csv)

</details>

## Key Points

- TODO: EPA's key points, extracted verbatim (see `R/gen_narrative.R` /
  `narrative_generated.md`) and cross-checked against the live snapshot page.

## Background

TODO: EPA's background prose, extracted verbatim.

## About the Indicator

TODO: EPA's "about the indicator" prose, extracted verbatim.

## About the Data

### Indicator Notes

TODO: EPA's indicator notes, extracted verbatim.

### Data Sources

TODO: EPA's data sources paragraph, extracted verbatim. Cross-check every URL
against the live page — heat-related-deaths' archived docx had two CDC URLs
that had since changed on the published page.

## Technical Documentation

```{r}
#| label: td-link
#| echo: false
#| output: asis
cat(sprintf("- [Download related technical information (PDF)](%s)\n", meta$indicator$technical_documentation))
```

## References

<!--
TODO: EPA's numbered reference list, extracted via R/gen_narrative.R as an
<ol class="references"> with id="ref-N" anchors, so body superscripts like
"^9,10^" link to them. See heat-related-deaths' index.qmd for the exact
output shape gen_narrative.R produces.
-->

::: {.callout-note appearance="simple" icon=false}
Text on this page is EPA's own published wording (a U.S. Government work, not
subject to domestic copyright), extracted from EPA's source document and
verified paragraph by paragraph against the published page. See
`R/gen_narrative.R` and `NOTICE.md`.
:::
